Modal Abstractions for Virtualizing Memory Addresses

### 2024-11-15 examples.tex
compress some text enough that only the conclusions are over the page limit

%\begin{comment}
%\todo[inline]{Identity mappings are difficult, and our current approach won't quite work. Consider trying to have a virtual pointsto for an actual page table entry (i.e., that one could use to update a page table mapping), while also having a virtual pointsto for an address that entry mapped. With the current (let's call it v1) solution, we can't actually have both of those simultaneously!  That's because the PTE pointsto will assert full ownership of the physical memory cell holding the PTE as its data value, while the virtual pointsto for the data mapped by that entry will \emph{also} assert (fractional) ownership of all entries a page table walk would traverse.
%}
%\todo[inline,color=violet]{This doesn't seem to cause issues with the mapping/unmapping examples, only with changing intermediate page table pointers. The mapping example requires a virtual pointsto for the blank PTE, and once filled in that ownership can be immediately split to create the 512 new virtual pointsto assertions for the newly mapped page. Conversely, for unmapping we'd assume ownership of all the relevant virtual pointsto assertions for the page we're unmapping, at which point we can (with a bit of work) show that they all correspond to the same L1 PTE, and extract the 512 fractional shares of that entry from the pointsto assertions.  But changing intermediate page tables, as one would do for coallescing or splitting a superpage while preserving the virtual-to-physical mappings, couldn't be done without some really complicated separating implication tricks.}
%\todo[inline,color=green]{One possible approach to resolving this, which we came up with in our Tuesday meeting, is to recognize that the current (v1) virtual points-to is too strong, because it really doesn't care about \emph{owning} those fractional resources, it only cares that \emph{something} ensures the correct page table walk exists. Iris has a ghost map resource where authoritative ownership of an individual key-value pair can be handled as a resource.  (Colin was using this in the filesystem cache.)
%We can use that mechanism to separate the virtual-to-physical translation from the physical memory involved (Kolanski and Klein may have done something similar for different reasons): (fractional) virtual points-to assertions can be defined in terms of (fractional) ownership of these authoritative ghost map entry assertions, plus sharing an invariant that the current installed page table respects all entries of the mapping. Unmapping collects the authoritative map kvpairs from collecting the assertions, and then can remove them from the ghost map and update the page tables. Critically, physical ownership of the page tables then lives in the invariant on the current page table, so some virtual pointsto assertions can refer to memory in those page tables.
%This still works with the modality, since that invariant is also semantically a predicate on a page table root.
%Let's call this v2.
%}
%\end{comment}
\subsection{Traversing Live Page Tables}
\label{sec:traversing}
We build up to the main task of mapping a new page after traversing page tables in software.
The mapping operation of Figure \ref{fig:mapping_code} assumes an operation \textsf{walkpgdir} which must traverse the page tables
in order to locate the address of the L1 entry to update --- possibly allocating tables for levels 3, 2, and 1 in the process,
installing them into levels 4, 3, and 2, along the way.
Traversing the page tables is itself challenging functionality to verify: loading the current table root from \lstinline|cr3| is straightforward
(a \lstinline|mov| instruction), however this produces the physical address of \lstinline|cr3|, not the virtual address the kernel code would use to access that memory.
This problem repeats at each level of the page table: assuming the code has \emph{somehow} read the appropriate L4 (or L3, or L2) entry, those entries again
yield physical addresses, not virtual.

@@

Note that none of the verification for this function assumes specific page table levels --- logical parameter \textsf{v} represents the level
of the entry passed as an argument, and this code
is used for all three level transitions when traversing page tables (4 to 3, 3 to 2, 2 to 1).
This comes into play with a subtlety of the specification of \lstinline|pte_get_next_table| that we will
revisit several times: \lstinline|pte_get_next_table|'s specification
assumes it is given a virtual \emph{vpte-pointsto}
% \footnote{A \emph{vpte-pointsto}
% $\vaddr\mapsto_{\textsf{vpte,qfrac}} \textsf{\paddr~\val}$ is a virtual points-to granting
% access to virtual address $\vaddr$, which is known to hold \textsf{\val}, but additionally exposing the physical address
% $\paddr$ that the virtual address translates to. This is commonly used, as the name suggests, when updating PTEs,
% where we require assurance that we are accessing the intended PTE.
% Its definition is just like the regular virtual points-to in Figure \ref{fig:virtualpointstosharing},
% except taking the physical address as a parameter rather than existentially quantifying it.
% }
(a virtual points-to exposing the underlying physical address; see Section \ref{sec:mapnew}) granting access to the specified entry,
but its postcondition does not yield new virtual points-to assertions!
Instead it merely computes the base address of the next table, and returns adequate capabilities (discussed in Section \ref{subsec:identitymappings})
for the \emph{caller} to construct a vpte-pointsto for the next table level (if this is not an L1 entry ---
the caller knows which level of the table this is for).
\looseness=-1

@@

\subsubsection{Identity Mappings}
\label{subsec:identitymappings}
It is typical for kernels to need to convert between physical and virtual addresses, in both directions.
Traversing the page tables in software is the simplest way to convert a virtual address to a physical address; this is the context we are working up to.
However, implementing this virtual-to-physical (V2P) translation in this way ironically requires physical-to-virtual (P2V) translation,
because the addresses stored in page table entries are physical, but memory accesses issued by the OS code use virtual addresses.
There is no universal way to convert physical addresses to virtual --- doing so relies on the kernel maintaining careful invariants or
additional data structures to enable P2V translation.

In practice, VMM operations are performance-critical for many workloads, so most kernels opt for using invariants to make P2V conversion very fast,
rather than maintaining yet another data structure.
Most kernels maintain an invariant on their page tables that the virtual address of any page used for a page table lives at a virtual address
whose value is \emph{a constant offset from the physical address} --- a practice sometimes referred to as \emph{identity mapping} 
(even though the physical-to-virtual translation
is typically not literally the identity function, but adding a non-zero constant offset).\footnote{Some kernels do this for all physical memory on the machine, simplifying interaction
with DMA devices.
On newer platforms like RISC-V, this sometimes truly is an identity mapping ---
x86-64 machines are forced into offsets by backwards compatibility with bootloaders that cannot access the full memory space of the
machine.
}

@@

\end{array}
\]
\vspace{-1em}
\caption{Global Address-Space Invariant in Figure \ref{fig:peraspaceinvariant} extended with a ghost map bookkeeping identity mappings }
  \label{fig:peraspaceinvariant_with_p2v_extension}
\end{figure*}

@@

If the entry is not set, \textsf{pte\_initialize} is called (Line \ref{line:call_to_pte_initialize} in Figure \ref{fig:calltopteinitialize}) 
for a physical page (utilizing the page-allocator's \textsf{kalloc} -- currently the only axiomatized call in the the proof of \textsf{pte\_initialize} 
(Line \ref{line:call_to_kalloc} in Figure \ref{pteinitializespec}). 
\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry+KERNBASE}  \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{\_} \ast \lnot(\textsf{entry\_present entry\_val} }_{\rtv}$
$\specline{\textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{entry entryv} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{entry})}{\textsf{level}} }$
;;void pte_initialize(pte_t *entry) { ... Stack frame setup elided    
callq  81 <kalloc> ;;allocate a full zeroed page for 512 8-byte entries(*\label{line:call_to_kalloc}*)
mov    rax,-0x10[rbp] ;; Store into 'local'
;;entry->pfn = PTE_ADDR_TO_PFN((uintptr_t) local);
mov    -0x10[rbp],rax; mov    -0x8[rbp],rdi; mov    rax,[rdi]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry+KERNBASE} \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr} \ast \lnot(\textsf{entry\_present entry\_val}) }$
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\mathsf{vpte,1}} \mathsf{entry} \; \textsf{pfn\_set(entryv nextpaddr)}}$
$\specline{\ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \begin{array}{l} \ulcorner\textsf{entry\_present (pte\_initialize(pfn\_set(entry\_val nextpaddr)),level)}\urcorner \wand \\  \;\;\;\;\;\;\; \forall_{i\in\textsf{0..511}} \; \textsf{table\_root (pte\_initialized (pfn\_set (entry\_val nextpaddr))) + i * 8} \mapsto_{\textsf{id}} \textsf{level-1}    \end{array}  }_{\rtv}$
... ;;clean up the stack, return
\end{lstlisting}
\vspace{-1em}
\caption{Allocating a physical page }
\label{pteinitializespec}
\vspace{-1em}
\end{figure}

@@

\subsubsection{Physical-to-Virtual Conversion with \textsf{P2V}}
\label{sec:p2v}
Once we reach to the certain knowledge of having an entry with a frame referencing to an allocated resource resides inside the identity mappings
(which can already be known if the branch at Line \ref{line:check_entry_present_jump} is taken, or ensured by allocating and installing a new entry
as just discussed for Lines \ref{line:check_entry_present_jump}--\ref{line:end_of_allocation_path}), 
we can utilize this knowledge 
to convert this frame address into an virtual address of the next page table through, again, identity mappings -- Line 57 in Figure \ref{fig:calltopteinitialize} 
specified in Figure \ref{fig:p2v}.

This actually constitutes a very critical piece of the full page table walk verification, and we have verified the critical step for a small x86-64 kernel, 
which is the physical-to-virtual conversion (often appearing as a macro \texttt{P2V} in C source code). 
In our small kernel (Line \ref{line:p2v} in Figure \ref{fig:p2v}), as in larger kernels, \texttt{P2V} is actually just addition by the
constant offset mentioned in Section \ref{subsec:identitymappings}, but the correctness 
of this simple instruction is quite subtle and relies on the extended invariant (Figure \ref{fig:peraspaceinvariant_with_p2v_extension})
explained in that section.

>>>>

%\begin{comment}
%\todo[inline]{Identity mappings are difficult, and our current approach won't quite work. Consider trying to have a virtual pointsto for an actual page table entry (i.e., that one could use to update a page table mapping), while also having a virtual pointsto for an address that entry mapped. With the current (let's call it v1) solution, we can't actually have both of those simultaneously!  That's because the PTE pointsto will assert full ownership of the physical memory cell holding the PTE as its data value, while the virtual pointsto for the data mapped by that entry will \emph{also} assert (fractional) ownership of all entries a page table walk would traverse.
%\todo[inline,color=violet]{This doesn't seem to cause issues with the mapping/unmapping examples, only with changing intermediate page table pointers. The mapping example requires a virtual pointsto for the blank PTE, and once filled in that ownership can be immediately split to create the 512 new virtual pointsto assertions for the newly mapped page. Conversely, for unmapping we'd assume ownership of all the relevant virtual pointsto assertions for the page we're unmapping, at which point we can (with a bit of work) show that they all correspond to the same L1 PTE, and extract the 512 fractional shares of that entry from the pointsto assertions.  But changing intermediate page tables, as one would do for coallescing or splitting a superpage while preserving the virtual-to-physical mappings, couldn't be done without some really complicated separating implication tricks.}
%\todo[inline,color=green]{One possible approach to resolving this, which we came up with in our Tuesday meeting, is to recognize that the current (v1) virtual points-to is too strong, because it really doesn't care about \emph{owning} those fractional resources, it only cares that \emph{something} ensures the correct page table walk exists. Iris has a ghost map resource where authoritative ownership of an individual key-value pair can be handled as a resource.  (Colin was using this in the filesystem cache.)
%We can use that mechanism to separate the virtual-to-physical translation from the physical memory involved (Kolanski and Klein may have done something similar for different reasons): (fractional) virtual points-to assertions can be defined in terms of (fractional) ownership of these authoritative ghost map entry assertions, plus sharing an invariant that the current installed page table respects all entries of the mapping. Unmapping collects the authoritative map kvpairs from collecting the assertions, and then can remove them from the ghost map and update the page tables. Critically, physical ownership of the page tables then lives in the invariant on the current page table, so some virtual pointsto assertions can refer to memory in those page tables.
%This still works with the modality, since that invariant is also semantically a predicate on a page table root.
%Let's call this v2.
%\end{comment}
\subsection{Traversing Live Page Tables}
\label{sec:traversing}
We build up to the main task of mapping a new page after traversing page tables in software.
The mapping operation of Figure \ref{fig:mapping_code} assumes an operation \textsf{walkpgdir} which must traverse the page tables
in order to locate the address of the L1 entry to update --- 
% possibly allocating tables for levels 3, 2, and 1 in the process,
% installing them into levels 4, 3, and 2, along the way.
possibly allocating new L3, L2, and L1 tables as necessary.
Traversing the page tables is itself challenging functionality to verify: loading the current table root from \lstinline|cr3| is straightforward
(a \lstinline|mov| instruction), however this produces the physical address of \lstinline|cr3|, not the virtual address the kernel code would use to access that memory.
This problem repeats at each level of the page table: assuming the code has \emph{somehow} read the appropriate L4 (or L3, or L2) entry, those entries again
yield physical addresses, not virtual.

@@

Note that the specification does \emph{not} assume a specific page table --- logical parameter \textsf{v} represents the level
of the entry passed as an argument, and this code
is used for all three level transitions when traversing page tables (4 to 3, 3 to 2, 2 to 1).
This comes into play with a subtlety of the specification of \lstinline|pte_get_next_table| that we will
revisit several times: \lstinline|pte_get_next_table|'s specification
assumes it is given a virtual \emph{vpte-pointsto}
% \footnote{A \emph{vpte-pointsto}
% $\vaddr\mapsto_{\textsf{vpte,qfrac}} \textsf{\paddr~\val}$ is a virtual points-to granting
% access to virtual address $\vaddr$, which is known to hold \textsf{\val}, but additionally exposing the physical address
% $\paddr$ that the virtual address translates to. This is commonly used, as the name suggests, when updating PTEs,
% where we require assurance that we are accessing the intended PTE.
% Its definition is just like the regular virtual points-to in Figure \ref{fig:virtualpointstosharing},
% except taking the physical address as a parameter rather than existentially quantifying it.
% }
(a virtual points-to exposing the underlying physical address; see Section \ref{sec:mapnew}) granting access to the specified entry,
but its postcondition does not yield new virtual points-to assertions!
Instead it merely computes the base address of the next table, and returns adequate capabilities (discussed in Section \ref{subsec:identitymappings})
for the \emph{caller} to construct a vpte-pointsto for the next table level (if this is not an L1 entry ---
the caller knows which level of the table this is for).
\looseness=-1

@@

\subsubsection{Identity Mappings}
\label{subsec:identitymappings}
Kernels need to convert between physical and virtual addresses, in both directions.
Traversing the page tables in software is the simplest way to convert a virtual address to a physical address; this is the context we are working up to.
However, implementing this virtual-to-physical (V2P) translation in this way ironically requires physical-to-virtual (P2V) translation,
because the addresses stored in page table entries are physical, but memory accesses issued by the OS code use virtual addresses.
% There is no universal way to convert physical addresses to virtual --- doing so relies on the kernel maintaining careful invariants or
% additional data structures to enable P2V translation.
\looseness=-1

Because VMM operations are performance-critical for many workloads, most kernels 
maintain invariants that enable very fast P2V conversions (rather than adding another data structure).
Most kernels maintain an invariant on their page tables that the virtual address of any page used for a page table lives at a virtual address
whose value is \emph{a constant offset from the physical address} --- a practice sometimes referred to as \emph{identity mapping} 
(even though the physical-to-virtual translation
is typically not literally the identity function, but adding a non-zero constant offset).\footnote{Some kernels do this for all physical memory on the machine, simplifying interaction
with DMA devices.
On newer platforms like RISC-V, this sometimes truly is an identity mapping ---
x86-64 machines are forced into offsets by backwards compatibility with bootloaders that cannot access the full memory space of the
machine.
}

@@

\end{array}
\]
\vspace{-1em}
\caption{Global Address-Space Invariant in Figure \ref{fig:peraspaceinvariant} extended with a ghost map bookkeeping identity mappings }
  \label{fig:peraspaceinvariant_with_p2v_extension}
\vspace{-1em}
\end{figure*}

@@

If the entry is not set, \textsf{pte\_initialize} (Line \ref{line:call_to_pte_initialize} in Figure \ref{fig:calltopteinitialize}) 
allocates a physical page (internally utilizing the only axiomatized call in our case studies, the page-allocator's \textsf{kalloc}, 
on Line \ref{line:call_to_kalloc} in Figure \ref{pteinitializespec}). 
\looseness=-1
\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry+KERNBASE}  \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{\_} \ast \lnot(\textsf{entry\_present entry\_val} }_{\rtv}$
$\specline{\textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{entry entryv} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{entry})}{\textsf{level}} }$
;;void pte_initialize(pte_t *entry) { ... Stack frame setup elided    
callq  81 <kalloc> ;;allocate a full zeroed page for 512 8-byte entries(*\label{line:call_to_kalloc}*)
mov    rax,-0x10[rbp] ;; Store into 'local'
;;entry->pfn = PTE_ADDR_TO_PFN((uintptr_t) local);
mov    -0x10[rbp],rax; mov    -0x8[rbp],rdi; mov    rax,[rdi]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry+KERNBASE} \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr} \ast \lnot(\textsf{entry\_present entry\_val}) }$
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\mathsf{vpte,1}} \mathsf{entry} \; \textsf{pfn\_set(entryv nextpaddr)}}$
$\specline{\ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \begin{array}{l} \ulcorner\textsf{entry\_present (pte\_initialize(pfn\_set(entry\_val nextpaddr)),level)}\urcorner \wand \\  \;\;\;\;\;\;\; \forall_{i\in\textsf{0..511}} \; \textsf{table\_root (pte\_initialized (pfn\_set (entry\_val nextpaddr))) + i * 8} \mapsto_{\textsf{id}} \textsf{level-1}    \end{array}  }_{\rtv}$
... ;;clean up the stack, return
\end{lstlisting}
\vspace{-1em}
\caption{Allocating a physical page }
\label{pteinitializespec}
\vspace{-1em}
\end{figure}

@@

\subsubsection{Physical-to-Virtual Conversion with \textsf{P2V}}
\label{sec:p2v}
Once we know the entry refers to an allocated resource inside the identity mapping range ($\Xi$)
% Once we reach to the certain knowledge of having an entry with a frame referencing to an allocated resource resides inside the identity mappings
(via the branch at Line \ref{line:check_entry_present_jump}, or  by allocating and installing a new entry
as just discussed for Lines \ref{line:check_entry_present_jump}--\ref{line:end_of_allocation_path}), 
we can %utilize this knowledge 
to convert this frame address into an virtual address of the next page table through, again, identity mappings -- Line 57 in Figure \ref{fig:calltopteinitialize} 
specified in Figure \ref{fig:p2v}.
This actually constitutes a very critical piece of the full page table walk verification, and we have verified the critical step for a small x86-64 kernel, 
which is the physical-to-virtual conversion (often appearing as a macro \texttt{P2V} in C source code). 
In our small kernel (Line \ref{line:p2v} in Figure \ref{fig:p2v}), as in larger kernels, \texttt{P2V} is actually just addition by the
constant offset mentioned in Section \ref{subsec:identitymappings}, but the correctness 
of this simple instruction is quite subtle and relies on the extended invariant (Figure \ref{fig:peraspaceinvariant_with_p2v_extension})
explained in that section.
